Data Prep: IUCN and AquaMaps

### 2016-12-19 data_explore/data_explore.Rmd
revisiting analysis based on new matching algorithm

message('Getting IUCN maps list')
iucn_maps_file <- file.path(dir_data, 'spp_iucn_maps.csv')
spp_iucn_maps <- read_csv(iucn_maps_file) %>%
  filter(spp_group != 'hagfishes') %>%           ### not from IUCN...
  filter(sciname %in% spp_iucn$sciname)   ### ditch freshwater spp

@@

sciname_lookup <- read_csv(file.path(dir_data, 'sciname_lookups/spp_iucn_scinames.csv'))

@@

iucn_maps_class_all <- spp_iucn_maps %>% 
  left_join(sciname_lookup %>% 
              dplyr::select(sciname, class) %>%
              unique(), 
            by = 'sciname') %>%
  select(-iucn_subpop, -spp_group) %>%
  distinct() %>% ### first, get rid of dupes due to subpopulations
  mutate(class    = str_to_title(class),
         am_match = sciname %in% spp_am$sciname)

@@

taxa_omitted <- taxa_am_all %>%
  full_join(taxa_am_solo,   by = 'class') %>%
  full_join(taxa_iucn_all,  by = 'class') %>%
  full_join(taxa_iucn_solo, by = 'class') %>%
  full_join(taxa_both,      by = 'class') %>%
  filter(is.na(n_iucn_class_all) & p_am_class_all < 0.5)

@@

taxa_df2 <- taxa_all_df %>%
  bind_rows(taxa_all_df %>%
              mutate(class = ' ',
                     start = 0,
                     end   = 0,
                     n_total = NA)) %>%
  bind_rows(taxa_df1 %>%
              select(-IUCN, -AquaMaps, -Paired) %>%
              arrange(desc(n_total))) %>%
  bind_rows(taxa_omitted %>%
              summarize(class = 'other (47 classes)',
                        n_total = sum(n_am_class_all),
                        dataset = 'AquaMaps',
                        start = 0,
                        end = 100)) %>%
  mutate(class = factor(class, levels = unique(.$class), ordered = TRUE),
         n_label = ifelse(!is.na(n_total), paste0('n = ', n_total), ''),
         dataset = factor(dataset, levels = c('AquaMaps', 'Paired', 'IUCN'), ordered = TRUE))

@@

bar_taxa <- ggplot(taxa_df2, aes(x = class, color = dataset)) +
  ggtheme_plot + 
  theme(panel.grid.major.x = element_blank()) +
  geom_linerange(aes(ymin = start, ymax = end, size = dataset),
                 position = position_dodge(width = 0.5)) +
  scale_size_manual(values = c('IUCN' = .5, 'AquaMaps' = .5, 'Paired' = 2.8)) +
  scale_color_manual(values = c('IUCN' = 'black', 'AquaMaps' = 'grey70', 'Paired' = 'black')) +
  geom_text(aes(label = n_label, y = 102),
            size = 2.4,
            angle = 90, hjust = 0, color = 'grey30') +
  theme(axis.text.x = element_text(angle = 75, hjust = 1),
        axis.title.y = element_text(hjust = 0)) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 140),
                     breaks = break_nums,
                     labels = sprintf('%s%%', break_nums)) +
  labs(x = 'Class (in order of decreasing number of included species)',
       y = '% of mapped species',
       fill = 'Dataset')

>>>>

message('Getting IUCN maps list')
iucn_maps_file <- file.path(dir_data, 'spp_iucn_maps.csv')
spp_iucn_maps <- read_csv(iucn_maps_file)

@@

@@

iucn_maps_class_all <- spp_iucn_maps %>% 
  select(-iucn_subpop, -spp_group) %>%
  distinct() %>% ### first, get rid of dupes due to subpopulations
  mutate(class    = str_to_title(class),
         am_match = sciname %in% spp_am$sciname)

@@

taxa_omitted <- taxa_am_all %>%
  full_join(taxa_am_solo,   by = 'class') %>%
  full_join(taxa_iucn_all,  by = 'class') %>%
  full_join(taxa_iucn_solo, by = 'class') %>%
  full_join(taxa_both,      by = 'class') %>%
  filter(is.na(n_iucn_class_all) & p_am_class_all < 0.5) %>%
  left_join(read_csv(file.path(dir_data, 'sciname_lookups/scinames.csv')) %>%
              select(class, kingdom, phylum) %>%
              mutate(class = str_to_title(class)),
            by = 'class') %>%
  mutate(class = paste('other', str_to_title(kingdom), sep = ' ')) %>%
  select(-kingdom, -phylum) %>%
  distinct()

@@

taxa_df2 <- taxa_all_df %>%
  bind_rows(taxa_all_df %>%
              mutate(class = ' ',
                     start = 0,
                     end   = 0,
                     n_total = NA)) %>%
  bind_rows(taxa_df1 %>%
              select(-IUCN, -AquaMaps, -Paired) %>%
              arrange(desc(n_total))) %>%
  bind_rows(x <- taxa_omitted %>%
              group_by(class) %>%
              summarize(n_total = sum(n_am_class_all),
                        dataset = 'AquaMaps',
                        start = 0,
                        end = 100)) %>%
  mutate(class = factor(class, levels = unique(.$class), ordered = TRUE),
         n_label = ifelse(!is.na(n_total), paste0('n = ', n_total), ''),
         dataset = factor(dataset, levels = c('AquaMaps', 'Paired', 'IUCN'), ordered = TRUE))

@@

bar_taxa <- ggplot(taxa_df2, aes(x = class, color = dataset)) +
  ggtheme_plot + 
  theme(panel.grid.major.x = element_blank()) +
  geom_linerange(aes(ymin = start, ymax = end, size = dataset),
                 position = position_dodge(width = 0.5)) +
  scale_size_manual(values = c('IUCN' = .5, 'AquaMaps' = .5, 'Paired' = 2.6)) +
  scale_color_manual(values = c('IUCN' = 'black', 'AquaMaps' = 'grey70', 'Paired' = 'black')) +
  geom_text(aes(label = n_label, y = 102),
            size = 2.4,
            angle = 90, hjust = 0, color = 'grey30') +
  theme(axis.text.x = element_text(angle = 75, hjust = 1),
        axis.title.y = element_text(hjust = 0)) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 140),
                     breaks = break_nums,
                     labels = sprintf('%s%%', break_nums)) +
  labs(x = 'Class (in order of decreasing number of included species)',
       y = '% of mapped species',
       fill = 'Dataset')

### 2016-12-19 data_explore/data_prep.Rmd
revisiting analysis based on new matching algorithm

dir_N <- c('Windows' = '//neptune.nceas.ucsb.edu/data_edit',
           'Darwin'  = '/Volumes/data_edit',
           'Linux'   = '/var/data/ohi')[[ Sys.info()[['sysname']] ]]

@@

dir_anx <- file.path(dir_N, 'git-annex/globalprep/SPP_ICO')
dir_fig <- file.path(dir_git, 'figures')
dir_data <- file.path(dir_git, 'data')
dir_data_remote <- file.path(dir_N, 'git-annex/globalprep/_raw_data')
scenario <- 'v2016'

@@

Need to:

gather spp_all, spp_am, spp_iucn, spp_iucn_maps, and spp_am_raw into local data folder.

``` {r pull_files_to_local, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}
### pull spp_iucn_marine.csv from git-annex and bring to local repo
spp_iucn <- read_csv(file.path(dir_anx, scenario, 'int/spp_iucn_cleaned.csv'))
spp_iucn <- spp_iucn %>%
  dplyr::select(sciname, iucn_sid, iucn_category, 
                popn_trend, parent_sid, subpop_sid) %>%
  unique()
write_csv(spp_iucn, file.path(dir_data, 'spp_iucn.csv'))

### pull spp_iucn_maps_all.csv from git-annex and bring to local repo
spp_iucn_maps <- read_csv(file.path(dir_anx, scenario, 'int/spp_iucn_maps_all.csv')) %>%
  rename(iucn_sid = id_no)

spp_iucn_maps <- spp_iucn_maps %>%
  filter(!str_detect(spp_group, 'MARINEFISH')) %>%
  unique() %>%
  arrange(iucn_sid)

### At this point, still some duplicated iucn_sids, due to several
### names matching a single ID.  Not sure which names might be in
### AquaMaps though, so leave 'em in.

write_csv(spp_iucn_maps, file.path(dir_data, 'spp_iucn_maps.csv'))

### pull cleaned spp_am data from git-annex and bring to local repo
spp_am <- read_csv(file.path(dir_anx, scenario, 'int/spp_am_cleaned.csv'))
write_csv(spp_am, file.path(dir_data, 'spp_am.csv'))

### pull raw aquamaps species list from git-annex and bring to local repo
spp_am_raw <- read_csv(file.path(dir_data_remote, 'aquamaps/v2015/csv/speciesoccursum.csv'))
write_csv(spp_am_raw, file.path(dir_data, 'speciesoccursum.csv'))

@@

```{r get_spp_all, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}
### get spp_all_complete from git-annex, bring to local repo.  The spp_all_complete
### file was created in ohiprep/globalprep/SPP_ICO/data_prep_SPP.R, running the
### function create_spp_master_lookup(source_pref = 'iucn', fn_tag = '_complete', reload = TRUE)
spp_all_remote <- read_csv(file.path(dir_anx, scenario, 'int/spp_all.csv'))
spp_all_remote_clean <- spp_all_remote %>%
  filter(!str_detect(spatial_source, 'subpop')) %>%
  filter(!str_detect(spatial_source,  'alias'))
### At this point, plenty of duplicates:
### * This code returns 984 duplicated names:
###     length(spp_all_remote_clean$sciname) - length(unique(spp_all_remote_clean$sciname))
### * some are AM subpops which were ignored in processing the remote spp_all file.
###   Note: All subpops have an identified parent, so can filter them out without
###   losing any species.
###   code to verify this:
###     z_parent <- spp_all_remote_clean %>% 
###       filter(!is.na(subpop_sid) & spatial_source == 'am')
###     z_subpop <- spp_all_remote_clean %>% 
###       filter(!is.na(parent_sid) & spatial_source == 'am')
###     nrow(z_subpop %>% filter(!sciname %in% z_parent$sciname))
### * some are duplicated IUCN parents (e.g. Dermochelys coriacea) that will be 
###   eliminated when select(-iucn_subpop) column and then unique()

### filter out spatial_source == 'am' that also have parent_sid (i.e.
### these are AM-mapped subpopulations); then: 
### select away the iucn_subpop, parent_sid, and subpop_sid (among others),
### these are creating false unique rows.
spp_all <- spp_all_remote_clean %>%
  filter(!(spatial_source == 'am' & !is.na(parent_sid))) %>%
  dplyr::select(am_sid, sciname, reviewed, iucn_sid, popn_trend,
                popn_category, category_score, spp_group, 
                id_no, spatial_source, name_verified) %>%
  unique()
### At this point, still some duped scinames - these are ones with multiple IUCN
### SIDs that are not designated as parents or subpops.  We will brute-force
### this by just eliminating the duplicated rows, keeping only the first instance.
### By arranging by iucn_sid, we will keep the lowest value, which seems to
### generally correspond to the parent...
spp_all <- spp_all %>%
  arrange(iucn_sid)
spp_all <- spp_all[!duplicated(spp_all$sciname), ]

length(spp_all$sciname) - length(unique(spp_all_remote_clean$sciname))
### no duped scinames - same length as original (cleaned) list

### But, some of the extracted IUCN maps don't appear on this list! add 'em in.
spp_iucn_maps_new <- read_csv(file.path(dir_data, 'spp_iucn_maps.csv')) %>%
  filter(!sciname %in% spp_all$sciname) %>%
  filter(!str_detect(sciname, 'subpop')) %>% ### a Hawaiian subpop... ditch it!
  select(-iucn_subpop)

spp_all1 <- spp_all %>%
  full_join(spp_iucn_maps_new)

### Also, some of the AM listings have disappeared due to name conflicts
### - aliases, etc.
spp_am_new <- read_csv(file.path(dir_data, 'speciesoccursum.csv')) %>%
  mutate(sciname = paste(str_trim(genus), str_trim(species)),
         iucn_sid = as.integer(iucn_id)) %>%
  dplyr::select(am_sid = speciesid, sciname, reviewed, iucn_sid,
                popn_category = iucn_code) %>%
  filter(!am_sid %in% spp_all$am_sid) %>%
  mutate(spatial_source = 'am',
         popn_category = ifelse(popn_category == 'LR/lc', 'LC', popn_category),
         popn_category = ifelse(popn_category == 'LR/nt', 'NT', popn_category),
         popn_category = ifelse(popn_category == 'N.E.',  'NE', popn_category))

### 545 new spp_am not in spp_all from remote source, by am_sid.
### 24 of these, however, list iucn_sids that match existing iucn_sid.
### * for species with iucn_sid that match, and no am_sid, add am_sid to 
###   existing record (use existing values for other fields; drop new values).
### * for species with iucn_sid that match, and with a different am_sid, no action.
### * for species with non-matching or NA iucn_sid, just add to list.
spp_am_new_iucnmatch <- spp_am_new %>% 
  filter(!is.na(iucn_sid) & iucn_sid %in% spp_all$iucn_sid) %>%
  dplyr::select(iucn_sid, am_sid_new = am_sid, reviewed_new = reviewed)
spp_all2 <- spp_all1 %>%
  left_join(spp_am_new_iucnmatch, by = 'iucn_sid') %>%
  mutate(am_sid   = ifelse(is.na(am_sid) & !is.na(am_sid_new), am_sid_new, am_sid),
         reviewed = ifelse(is.na(am_sid) & !is.na(am_sid_new), reviewed_new, reviewed)) %>%
  select(-am_sid_new, -reviewed_new)
spp_all3 <- spp_all2 %>%
  full_join(spp_am_new %>%
              filter(!iucn_sid %in% spp_am_new_iucnmatch$iucn_sid),
            by = c("am_sid", "sciname", "reviewed", "iucn_sid", "popn_category", "spatial_source"))

write_csv(spp_all3, file.path(dir_data, 'spp_all.csv'))

@@

```{r get_other_vars, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}
### create spp_am_v_iucn: the whole maplist narrowed down to just the overlapping species.
### Overlaps detected from spatial_source containing 'iucn' (b/c source_pref = 'iucn' above),
### that also have a non-NA AquaMaps ID code.
spp_all <- read_csv(file.path(dir_data, 'spp_all.csv'))

spp_am_v_iucn <- spp_all %>%
  filter(str_detect(spatial_source, 'iucn') & !is.na(am_sid)) %>%
  filter(!is.na(id_no)) %>%
  filter(!str_detect(spp_group, 'MARINEFISH')) %>%
  unique()
# length(unique(spp_am_v_iucn$am_sid)); length(unique(spp_am_v_iucn$iucn_sid))
### NOTE: 2345 unique scientific names; but 2283 unique am_sid and 2249 unique iucn_sid.
### Some unique species (as identified by am_sid and iucn_sid) have multiple
### scinames.  At this point, not really important what the sciname is?
spp_am_v_iucn1 <- spp_am_v_iucn %>%
  group_by(am_sid, iucn_sid) %>%
  mutate(dupes = n() - 1,
         first_instance = sciname == first(sciname)) %>%
  filter(!(dupes > 0 & !first_instance)) %>%
  select(-dupes, -first_instance) %>%
  ungroup()
# length(unique(spp_am_v_iucn1$am_sid)); length(unique(spp_am_v_iucn1$iucn_sid))

### What species are represented by the duplicated IUCN ID numbers?  These
### are species that IUCN experts lumped together (via synonyms etc) 
### but are treated as separate by AquaMaps (the reverse does not seem
### to happen)...  There are 64 AM species IDs and only 30 IUCN species IDs.  
### How to treat in the mapping?
### * lump AM species together and plot against a single IUCN species map?
###   That would occur in the area determining function - group by iucn_sid
###   and map all am_sids against that iucn_sid
x <- spp_am_v_iucn1 %>%
  group_by(iucn_sid) %>%
  filter(n() > 1)
# length(unique(x$am_sid)); length(unique(x$iucn_sid))

write_csv(spp_am_v_iucn1, file.path(dir_data, 'spp_am_v_iucn.csv'))
```

@@

### This file is generated in globalprep/SPP_ICO:
iucn_spp_cells_all <- read_csv(file.path(dir_data_remote, 'iucn_spp/iucn_cells_v2016.csv'),
                            col_types = 'cddddc')

iucn_spp_cells <- iucn_spp_cells_all %>%
  rename(iucn_sid = id_no) %>%
  filter(iucn_sid %in% spp_am_v_iucn$iucn_sid) %>%
  filter(presence != 5)
### length(unique(iucn_spp_cells$iucn_sid)) results in 2279 species found...
### some species with no cells reported.  2279 listed in the am v iucn file.
write_csv(iucn_spp_cells, file.path(dir_anx, 'vAM_IUCN/iucn_spp_cells_2015.csv'))

message('Loading AquaMaps species to cell table.  This might take a few minutes.')
am_spp_cells_all <- fread(file.path(dir_data_remote, 'aquamaps/v2015/csv/hcaf_sp_native_trunc.csv'), 
                         showProgress = TRUE)
am_spp_cells <- am_spp_cells_all %>%
  rename(am_sid = speciesid, prob = probability) %>%
  filter(am_sid %in% spp_am_v_iucn$am_sid)
### length(unique(am_spp_cells$am_sid)) results in 2314 species found - great!
write_csv(am_spp_cells, file.path(dir_anx, 'vAM_IUCN/am_spp_cells_2015.csv'))

@@

  if(!exists('am_spp_cells'))
    am_spp_cells <- read_csv(file.path(dir_anx, 'vAM_IUCN/am_spp_cells_2015.csv'))
  if(!exists('iucn_spp_cells'))
    iucn_spp_cells <- read_csv(file.path(dir_anx, 'vAM_IUCN/iucn_spp_cells_2015.csv'),
                               col_types = 'cddddc')
  ### species info list - scinames, IUCN species ID, AM species ID, category etc
  spp_list_base <- read_csv(file.path(dir_data, 'spp_am_v_iucn.csv'))

>>>>

dir_M <- c('Windows' = '//mazu.nceas.ucsb.edu/data_edit',
           'Darwin'  = '/Volumes/ohi',
           'Linux'   = '/var/data/ohi')[[ Sys.info()[['sysname']] ]]

@@

dir_anx <- file.path(dir_M, 'git-annex/globalprep/spp_ico')
dir_fig <- file.path(dir_git, 'figures')
dir_data <- file.path(dir_git, 'data')
dir_data_remote <- file.path(dir_M, 'git-annex/globalprep/_raw_data')
scenario <- 'v2016'

@@

```{r get_spp_all, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

spp_iucn_maps <- read_csv(file.path(dir_data, 'spp_iucn_maps.csv'))

spp_am_maps <- read_csv(file.path(dir_data, 'speciesoccursum.csv')) %>%
  mutate(sciname = paste(str_trim(genus), str_trim(species)),
         iucn_id = as.integer(iucn_id)) %>%
  dplyr::select(am_sid = speciesid, sciname, reviewed, iucn_id)

spp_all <- spp_iucn_maps %>%
  full_join(spp_am_maps, by = 'sciname') %>%
  distinct()

### some confusion due to non-matching scinames and IDs...
spp_all <- spp_all %>%
  mutate(iucn_match = iucn_sid == iucn_id,
         iucn_wrong_map = ifelse(!iucn_match & iucn_id %in% spp_all$iucn_sid, TRUE, FALSE)) %>%
  mutate(iucn_sid = ifelse(iucn_wrong_map & !is.na(iucn_wrong_map), iucn_id, iucn_sid)) %>%
  select(-iucn_match, -iucn_wrong_map, -iucn_id, -sciname_count) %>%
  distinct()

length(unique(x$iucn_sid)) ###  4013, incl NA
length(unique(x$am_sid))   ### 22890, incl NA

write_csv(spp_all, file.path(dir_data, 'spp_all_maps.csv'))

@@

```{r get_other_vars, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}
### create spp_am_v_iucn: the whole maplist narrowed down to just the overlapping species.
### Overlaps detected from spatial_source containing 'iucn' (b/c source_pref = 'iucn' above),
### that also have a non-NA AquaMaps ID code.
spp_all <- read_csv(file.path(dir_data, 'spp_all_maps.csv'))

spp_am_v_iucn <- spp_all %>%
  select(-iucn_subpop) %>%
  filter(!is.na(iucn_sid) & !is.na(am_sid)) %>%
  unique()

# length(unique(spp_am_v_iucn$sciname)); length(unique(spp_am_v_iucn$am_sid)); length(unique(spp_am_v_iucn$iucn_sid))
### NOTE: 2314 unique scientific names; but 2314 unique am_sid and 2320 unique iucn_sid.
### Seasnakes are included in Reptiles (IUCN).  Two "mangroves" that are actually seagrasses.
spp_am_v_iucn <- spp_am_v_iucn %>%
  group_by(am_sid, iucn_sid) %>%
  mutate(dupes = n() - 1) %>%
  ungroup() %>%
  filter(!(dupes == 1 & spp_group == 'MANGROVES')) %>%
  filter(spp_group != 'SEASNAKES') %>%
  select(-dupes)
# length(unique(spp_am_v_iucn$sciname)); length(unique(spp_am_v_iucn$am_sid)); length(unique(spp_am_v_iucn$iucn_sid))

sciname_dupes <- spp_am_v_iucn %>%
  group_by(sciname) %>%
  filter(n() > 1)
### these seem to be spp with IUCN maps for synonyms.  Keep 'em in.

write_csv(spp_am_v_iucn, file.path(dir_data, 'spp_am_v_iucn.csv'))

@@

@@

# iucn_spp_cells_files <- list.files(file.path(dir_anx, 'v2016/iucn_intersections'), full.names = TRUE)
# iucn_spp_cells_all <- lapply(iucn_spp_cells_files, FUN = function(x) {
#   read_csv(x) %>% 
#     mutate(id_no = as.integer(id_no)) %>%
#     rename(iucn_sid = id_no)
#   }) %>%
#   bind_rows()
# 
# iucn_spp_cells <- iucn_spp_cells_all %>%
#   filter(iucn_sid %in% spp_am_v_iucn$iucn_sid) %>%
#   filter(presence != 5)

### length(unique(iucn_spp_cells$iucn_sid)) results in 2271 species found...
###   some species with no cells reported.  
### spp_missing <- spp_am_v_iucn %>% filter(!iucn_sid %in% iucn_spp_cells$iucn_sid)
###   49 spp with no cells reported.  Shapefile errors?
# write_csv(iucn_spp_cells, file.path(dir_anx, 'vAM_IUCN/iucn_spp_cells_2015.csv'))

message('Loading AquaMaps species to cell table.  This might take a few minutes.')
am_spp_cells_all <- fread(file.path(dir_data_remote, 'aquamaps/d2015/csv/hcaf_sp_native_trunc.csv'), 
                         showProgress = TRUE)
am_spp_cells <- am_spp_cells_all %>%
  rename(am_sid = speciesid, prob = probability) %>%
  filter(am_sid %in% spp_am_v_iucn$am_sid)
### length(unique(am_spp_cells$am_sid)) results in 2314 species found
write_csv(am_spp_cells, file.path(dir_anx, 'vAM_IUCN/am_spp_cells_2015.csv'))

@@

  if(!exists('am_spp_cells'))
    am_spp_cells <- read_csv(file.path(dir_anx, 'vAM_IUCN/am_spp_cells_2015.csv'))
  if(!exists('iucn_spp_cells'))
    iucn_spp_cells <- read_csv(file.path(dir_anx, 'vAM_IUCN/iucn_spp_cells_2015.csv'))

  ### species info list - scinames, IUCN species ID, AM species ID, category etc
  spp_list_base <- read_csv(file.path(dir_data, 'spp_am_v_iucn.csv'))
